### 2015-06-30 Paper_v1.tex
Working on the introduction

% Begin ...
\begin{document}
\begin{titlepage}
{\par\centering\textbf{\Large {Dynamically dimensioned particle swarm optimization (DdPSO): A meta-heuristic for parameter estimation in large biochemical models.}}}
\vspace{0.05in}
{\par \centering \large{Adithya Sagar, Christine Shoemaker and Jeffrey D. Varner$^{*}$}}
\vspace{0.10in}
{\par \centering \large{School of Chemical and Biomolecular Engineering}}
{\par \centering \large{Cornell University, Ithaca NY 14853}}
\vspace{0.1in}
{\par \centering \textbf{Running Title:}~Parameter estimation in biochemical models}
\vspace{0.1in}
{\par \centering \textbf{To be submitted:}~\emph{Biotechnology Journal}}
\vspace{0.5in}
{\par \centering $^{*}$Corresponding author:}
{\par \centering Jeffrey D. Varner,}
{\par \centering Associate Professor, School of Chemical and Biomolecular Engineering,}
{\par \centering 244 Olin Hall, Cornell University, Ithaca NY, 14853}
{\par \centering Email: jdv27@cornell.edu}
{\par \centering Phone: (607) 255 - 4258}
{\par \centering Fax: (607) 255 - 9166}
\end{titlepage}
\date{}
\thispagestyle{empty}
\pagebreak
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Abstract}
The problem of parameter estimation presents a significant challenge in the modeling of large complex systems. Mathematical modeling of biological systems with multiple feed back loops is one such area where parameter estimation is a difficult non-linear optimization problem. This difficulty is further compounded when dealing with parameter vectors of high dimensions. Such problems generally involve expensive function evaluations and it is hard to obtain optimal solutions within finite computational time. In this study we present a novel meta-heuristic approach that combines a variant of particle swarm optimization (PSO) along with dynamically dimensioned search (DDS) to obtain near optimal solutions of high dimensional biochemical networks within a relatively few function evaluations. We use a particle swarm optimization technique that uses multi-swarms to generate candidate vectors which are then greedily updated using DDS by dynamically varying the perturbed parameter dimensions. We tested this algorithm (25 trials with 4000 function evaluations in each trial) on a biochemical network of coagulation (148 parameters and 92 species) and compared it's performance against other meta-heuristics like Differential Evolution (DE), Particle Swarm Optimization (PSO), Simulated Annealing (SA) and also against DDS alone. The new algorithm outperforms all the other meta-heuristics on the coagulation model. The parameter vectors obtained using this approach fit the experimental data well and also make accurate enough predictions on unseen experimental data. We also performed this comparison on commonly used test functions (Ackley and Rosenbrock) for global optimization and found the same behavior. Further we used two recently published benchmark problems, a genome wide kinetic model with 1759 parameters and a metabolic model of Chinese Hamster Ovary cells with 117 parameters to evaluate the performance of our approach. We  surprisingly performed well on these benchmarks and obtained the nominal parameter vector with just 4000 function evaluations in both cases.

@@

Biological systems interact with one another and with the external world through highly complex biochemical networks. The advent of high throughout technologies has provided researchers with a deluge of data about these networks. Over the last decade, with an increase in computational power, large scale mathematical modeling has evolved as a powerful paradigm to characterize these complex biochemical interactions \cite{assmus2006dynamics, Riel:2006aa,Jaqaman:2006aa,kitano2002systems,hood2004systems}. Mathematical modeling of these networks is an iterative process that primarily consists of two tasks a) Model construction b) Model calibration or Parameter Estimation \cite{Aldridge:2006aa,banga2008optimization,ashyraliyev2009systems}. Model construction involves translating existing knowledge about the network (i.e. entities in the network, interactions amongst the entities, types of interactions) along with reasonable assumptions into an appropriate mathematical framework. Model calibration or parameter estimation follows the model construction step.  In this step the model is trained against experimental data to determine the model parameters that give a good 'fit' against experimental data. Model training is generally framed as a non-linear optimization problem wherein the residual (or objective function) between experimental data and simulation data is minimized using various optimization techniques \cite{moles2003parameter}. The optimal parameters obtained from the previous step are then used to validate the model on hitherto unseen experimental data. If validation leads to unsatisfactory results, model construction and calibration are repeated iteratively till satisfactory results are obtained.

With the capacity to construct very large models using various mathematical formalisms \cite{chen2009input,tasseff2011modeling,luan2007computationally,mo2007genome,orth2011comprehensive,karr2012whole,buchel2013path2models,smallbone2010towards}, more often than not, parameter estimation is becoming a major challenge if not a bottleneck in large scale mathematical modeling. Though parameter estimation has been a well studied problem in engineering \cite{nieman1971review,beck1977parameter,young1981parameter,beck1998inverse}, the complex dynamics of large biological systems and noisy, often incomplete experimental data pose a more difficult and rather unique challenge. Most of the optimization problems pertaining to biological systems are non-linear and multi-modal i.e. most of them have multiple local minima or maxima \cite{moles2003parameter,banga2008optimization}. The non-linearity of the problem coupled with multi-modality generally renders the local optimization techniques like pattern search method \cite{hooke1961direct}, Nelder-Mead simplex method \cite{nelder1965simplex} or Levenberg-Marquardt method\cite{more1978levenberg} incapable of obtaining an optimal solution. Though deterministic global optimization techniques can handle non-linearity and multi-modality \cite{esposito2000deterministic,horst2013global}, the absence of derivative information, discontinuity of the objective functions, non-smooth regions or the lack of any knowledge about the objective function severely hamper the use of these techniques.

Meta-heuristic approaches like Genetic Algorithms, Simulated Annealing \cite{kirkpatrick1983optimization}, Evolutionary Programming and population based searches like Differential Evolution\cite{storn1997differential} have shown promise in this regard \cite{sun2012parameter}. They do not make any assumptions about the structure of objective function or require any \textit{a priori} information about the objective function. Though they do not guarantee strong convergence, these approaches are effective in finding near optimal solutions.  Mendes et al. \cite{mendes1998non} used Simulated Annealing to estimate rate constants for the irreversible inhibition of HIV proteinase, Modchang et al. \cite {modchang2008mathematical} used Genetic Algorithms to estimate parameters for a mathematical model of signal transduction, Differential Evolution based approaches have been an effective approach on various systems \cite{tsai2005evolutionary,wang2001hybrid,noman2007inferring}. Tashkova et al. \cite{tashkova2011parameter} compared different meta-heuristics for parameter estimation on a dynamic model of endocytosis and showed that DE was the most effective. Banga and co-workers have applied scatter-search based methods \cite{villaverde2012cooperative,rodriguez2006novel,egea2007scatter} to estimate parameters on non-linear biological processes. Hybrid approaches that combine a meta-heuristic with a local optimization search, wherein a near optimal solution that is obtained with a meta-heuristic is further refined using a local search have also become popular. Villaverde et al. \cite{villaverde2015biopredyn} combined scatter search with local search methods for parameter estimation in large scale systems biology models. Fan et al. recently showed that population based meta-heuristics along with decomposition based methods can be used to model gene circuits from mRNA data \cite{fan2015parameter}.

Despite these successes, a  major drawback with most of these approaches is the vast number of objective function evaluations they take in arriving at good search regions. As the models grow in size and complexity and the number of dimensions of parameter vector increases,  parameter estimation becomes a computationally expensive task. In many of these high dimensional problems approaching an exact solution may not be necessary. Gutenkust et al. \cite{gutenkunst2007universally} showed that a number of systems biology models are 'sloppy'. Sloppy systems have specific parameter combinations that largely define the dynamics of the system. Large perturbations to the rest of the parameters does not greatly impact the system dynamics. Ensemble approaches \cite{song2010ensembles,luan2010ensembles} have exploited this aspect to describe the dynamics of biological systems including coagulation which can be described using only a set of key species or parameters \cite{sagar2015dynamic}. Tolson and Shoemaker \cite{tolson2007dynamically} showed through Dynamically Dimensioned Search (DDS) that high-dimensional watershed models can be calibrated quickly by perturbing only a subset of dimensions. Thus in our current study we took into cognizance the fact that it may not be necessary to obtain the exact solution for high dimensional biological systems and that good enough solutions can be quickly obtained without expending a lot of objective function evaluations. Hence our current approach uses the power of population based heuristics along with DDS to obtain near optimal or good solutions. Though in theory we can combine any meta-heuristic with DDS, for the current purpose we used Particle Swarm Optimization (PSO). PSO unlike Genetic Algorithm (GA) or Differential Evolution (DE) does not have complex operations like cross over, mutation or recombination. It is simple to use and does not have a lot of parameters associated with other heuristics. However PSO is known to rapidly converge to a local optimum and thus several variants \cite{peer2003using,zhan2009adaptive,li2007fast} have been developed of which Multi swarm particle swarm optimization approaches (MLSPSO) \cite{zhao2008dynamic,liang2005dynamic} are one of the effective ones. We used MLSPSO to preclude bad regions of search and thereafter search within these regions using DDS.

>>>>

% Begin ...
\begin{document}
\begin{titlepage}
{\par\centering\textbf{\Large {Dynamically dimensioned particle swarm optimization (DdPSO): A meta-heuristic for parameter estimation in biochemical models}}}
\vspace{0.05in}
{\par \centering \large{Adithya Sagar, Christine Shoemaker$^{\dag}$ and Jeffrey D. Varner$^{*}$}}
\vspace{0.10in}
{\par \centering {School of Chemical and Biomolecular Engineering}}
{\par \centering {$^{\dag}$School of Civil and Environmental Engineering}}
{\par \centering {Cornell University, Ithaca NY 14853}}
\vspace{0.1in}
{\par \centering \textbf{Running Title:}~Parameter estimation in biochemical models}
\vspace{0.1in}
{\par \centering \textbf{To be submitted:}~\emph{PLoS ONE}}
\vspace{0.5in}
{\par \centering $^{*}$Corresponding author:}
{\par \centering Jeffrey D. Varner,}
{\par \centering Associate Professor, School of Chemical and Biomolecular Engineering,}
{\par \centering 244 Olin Hall, Cornell University, Ithaca NY, 14853}
{\par \centering Email: jdv27@cornell.edu}
{\par \centering Phone: (607) 255 - 4258}
{\par \centering Fax: (607) 255 - 9166}
\end{titlepage}
\date{}
\thispagestyle{empty}
\pagebreak
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Abstract}
The problem of parameter estimation presents a significant challenge in the modeling of large complex systems. Mathematical modeling of biological systems with multiple feed back loops is one such area where parameter estimation is a difficult non-linear optimization problem. This difficulty is further compounded when dealing with parameter vectors of high dimensions. Such problems generally involve expensive function evaluations and it is hard to obtain optimal solutions within finite computational time. In this study we present a novel meta-heuristic approach that combines a variant of particle swarm optimization (PSO) along with dynamically dimensioned search (DDS) to obtain near optimal solutions of high dimensional biochemical networks within a relatively few function evaluations. We use a particle swarm optimization technique that uses multi-swarms to generate candidate vectors which are then greedily updated using DDS by dynamically varying the perturbed parameter dimensions. We tested this algorithm (25 trials with 4000 function evaluations in each trial) on a biochemical network of coagulation (148 parameters and 92 species) and compared it's performance against other meta-heuristics like Differential Evolution (DE), Particle Swarm Optimization (PSO), Simulated Annealing (SA) and also against DDS alone. The new algorithm outperforms all the other meta-heuristics on the coagulation model. The parameter vectors obtained using this approach fit the experimental data well and also make accurate enough predictions on unseen experimental data. We also performed this comparison on commonly used test functions (Ackley and Rosenbrock) for global optimization and found the same behavior. Further we used two recently published benchmark problems, a genome wide kinetic model with 1759 parameters and a metabolic model of Chinese Hamster Ovary cells with 117 parameters to evaluate the performance of our approach. We  surprisingly performed well on these benchmarks and obtained the nominal parameter vector with just 4000 function evaluations in both cases.

@@

Cells process nutrients and respond to changes in their environment using complex biochemical networks. These networks contain thousands of components
interconnected through nonlinear enzyme catalyzed reactions. Mathematical modeling has evolved as a powerful paradigm to characterize, and ultimately design these complex networks \cite{assmus2006dynamics, Riel:2006aa,Jaqaman:2006aa,kitano2002systems,hood2004systems}. Mathematical modeling of biochemical networks is often an iterative process.
First, models are formulated from biochemical knowledge, and then trained using using experimental data \cite{Aldridge:2006aa,banga2008optimization,ashyraliyev2009systems}.
Model training is typically framed as a non-linear optimization problem wherein the residual (or objective function) between experimental data and model simulations is minimized using an optimization strategy \cite{moles2003parameter}. The optimal parameters obtained from training are then used to validate the model on hitherto unseen experimental data. If validation fails, model construction and calibration are repeated iteratively until satisfactory results are obtained.

Parameter estimation is a major challenge in the development of large biochemical models.
Though parameter estimation has been a well studied problem in engineering for decades \cite{nieman1971review,beck1977parameter,young1981parameter,beck1998inverse},
the complex dynamics of large biological systems and noisy, often incomplete experimental data pose a unique estimation challenge.
Most of the optimization problems pertaining to biological systems are non-linear and multi-modal i.e. most of them have multiple local minima or maxima \cite{moles2003parameter,banga2008optimization}. The non-linearity of the problem coupled with multi-modality generally renders the local optimization techniques like pattern search method \cite{hooke1961direct}, Nelder-Mead simplex method \cite{nelder1965simplex} or Levenberg-Marquardt method\cite{more1978levenberg} incapable of obtaining an optimal solution. Though deterministic global optimization techniques can handle non-linearity and multi-modality \cite{esposito2000deterministic,horst2013global}, the absence of derivative information, discontinuity of the objective functions, non-smooth regions or the lack of any knowledge about the objective function severely hamper the use of these techniques.

Meta-heuristic approaches like Genetic Algorithms, Simulated Annealing \cite{kirkpatrick1983optimization}, Evolutionary Programming and population based searches like Differential Evolution\cite{storn1997differential} have shown promise in this regard \cite{sun2012parameter}. They do not make any assumptions about the structure of objective function or require any \textit{a priori} information about the objective function. Though they do not guarantee strong convergence, these approaches are effective in finding near optimal solutions.  Mendes et al. \cite{mendes1998non} used Simulated Annealing to estimate rate constants for the irreversible inhibition of HIV proteinase, Modchang et al. \cite {modchang2008mathematical} used Genetic Algorithms to estimate parameters for a mathematical model of signal transduction, Differential Evolution based approaches have been an effective approach on various systems \cite{tsai2005evolutionary,wang2001hybrid,noman2007inferring}. Tashkova et al. \cite{tashkova2011parameter} compared different meta-heuristics for parameter estimation on a dynamic model of endocytosis and showed that DE was the most effective. Banga and co-workers have applied scatter-search based methods \cite{villaverde2012cooperative,rodriguez2006novel,egea2007scatter} to estimate parameters on non-linear biological processes. Hybrid approaches that combine a meta-heuristic with a local optimization search, wherein a near optimal solution that is obtained with a meta-heuristic is further refined using a local search have also become popular. Villaverde et al. \cite{villaverde2015biopredyn} combined scatter search with local search methods for parameter estimation in large scale systems biology models. Fan et al. recently showed that population based meta-heuristics along with decomposition based methods can be used to model gene circuits from mRNA data \cite{fan2015parameter}.Despite these successes, a  major drawback with most of these approaches is the vast number of objective function evaluations they take in arriving at good search regions. As the models grow in size and complexity and the number of dimensions of parameter vector increases,  parameter estimation becomes a computationally expensive task. In many of these high dimensional problems approaching an exact solution may not be necessary. Gutenkust et al. \cite{gutenkunst2007universally} showed that a number of systems biology models are 'sloppy'. Sloppy systems have specific parameter combinations that largely define the dynamics of the system. Large perturbations to the rest of the parameters does not greatly impact the system dynamics. Ensemble approaches \cite{song2010ensembles,luan2010ensembles} have exploited this aspect to describe the dynamics of biological systems including coagulation which can be described using only a set of key species or parameters \cite{sagar2015dynamic}. Tolson and Shoemaker \cite{tolson2007dynamically} showed through Dynamically Dimensioned Search (DDS) that high-dimensional watershed models can be calibrated quickly by perturbing only a subset of dimensions.

In this study, we developed the dynamically dimensioned particle swarm optimization (DdPSO) approach, a novel meta-heurstic that combined the global search capability of multi-swarm particle swarm optimization with dynamic dimensioned search. The design objective of DdPSO was to obtain near optimal parameter estimates for large biochemical models within a relatively few function evaluations.
DdPSO uses a multi-swarm particle swarm optimization technique to generate candidate solution vectors which are then greedily updated using dynamic dimensioned search.
We tested the performance of DdPSO on commonly used test functions for global optimization (Ackley and Rosenbrock), published biochemical parameter estimation benchmark problems and
lastly on a model of the human coagulation cascade. DdPSO outperformed common meta-heuristic approaches like Differential Evolution (DE), Particle Swarm Optimization (PSO), Simulated Annealing (SA) and dynamically dimensioned search (DDS) on the benchmark and coagulation problems.

% Extra text from the introduction -
%With the capacity to construct very large models using various mathematical formalisms \cite{chen2009input,tasseff2011modeling,luan2007computationally,mo2007genome,orth2011comprehensive,karr2012whole,buchel2013path2models,smallbone2010towards}, more often than not,

%this algorithm (25 trials with 4000 function evaluations in each trial) on a biochemical network of coagulation (148 parameters and 92 species) and compared it's performance against other meta-heuristics  The new algorithm outperforms all the other meta-heuristics on the coagulation model. The parameter vectors obtained using this approach fit the experimental data well and also make accurate enough predictions on unseen experimental data. We also performed this comparison on  and found the same behavior. Further we used two recently published benchmark problems, a genome wide kinetic model with 1759 parameters and a metabolic model of Chinese Hamster Ovary cells with 117 parameters to evaluate the performance of our approach. We  surprisingly performed well on these benchmarks and obtained the nominal parameter vector with just 4000 function evaluations in both cases.
%took into cognizance the fact that it may not be necessary to obtain the exact solution for high dimensional biological systems and that good enough solutions can be quickly obtained without expending a lot of objective function evaluations. Hence our current approach uses the power of population based heuristics along with DDS to obtain near optimal or good solutions. Though in theory we can combine any meta-heuristic with DDS, for the current purpose we used Particle Swarm Optimization (PSO). PSO unlike Genetic Algorithm (GA) or Differential Evolution (DE) does not have complex operations like cross over, mutation or recombination. It is simple to use and does not have a lot of parameters associated with other heuristics. However PSO is known to rapidly converge to a local optimum and thus several variants \cite{peer2003using,zhan2009adaptive,li2007fast} have been developed of which Multi swarm particle swarm optimization approaches (MLSPSO) \cite{zhao2008dynamic,liang2005dynamic} are one of the effective ones. We used MLSPSO to preclude bad regions of search and thereafter search within these regions using DDS.
